---
title: "Umpire 2.0: Clinically Realistic Simulations"
author: "Kevin R. Coombes and Caitlin E. Coombes"
data: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Umpire 2.0}
  %\VignetteKeywords{Umpire, simulations, mixed type data, clinical data}
  %\VignetteDepends{Umpire}
  %\VignettePackage{Umpire}
  %\VignetteEngine{knitr::rmarkdown}
---

# Introduction

Version 2.0 of the Ultimate Microarray Prediction, Inference, and 
Reality Engine (Umpire) extends the functions of the Umpire 1.0 R 
package to allow researchers to simulate realistic, mixed-type, 
clinical data. Statisticians, computer scientists, and clinical 
informaticians who develop and improve methods to analyze clinical data 
from a variety of contexts (including clinical trials, population 
cohorts, and electronic medical record sources) recognize that it is 
difficult to evaluate methods on real data where "ground truth" is unknown. 
Frequently, they turn to simulations where the can control the 
underlying structure, which can result in simulations which are too 
simplistic to reflect complex clinical data realities. Clinical 
measurements on patients may be treated as independent, in spite of 
the elaborate correlation structures that arise in networks, pathways, 
organ systems, and syndromes in real biology. Further, the researcher 
finds limited tools at her disposal to facilitate simulation of binary, 
categorical, or mixed data at this representative level of biological 
complexity.


In this vignette, we describe a workflow with the <tt>Umpire</tt> package 
to simulate biologically realistic, mixed-type clinical data.

As usual, we start by loading the package:

```{r}
library(Umpire)
```

# Simulating Mixed-Type Clinical Data
Since we are going to run simulations, for reproducibility purposes, we should
set the seed of the random number generator.
```{r seed}
set.seed(84503)
```

## Model Subtypes and Survival
The simulation workflow begins by simulating complex, correlated, 
continuous data with known "ground truth" by instantiating a 
<tt>ClinicalEngine</tt>. We simulate 20 features and 4 clusters of 
unequal size. The ClinicalEngine generates subtypes (clusters) with 
known "ground truth" through an implementation of the Umpire 1.0 
<tt>CancerModel</tt> and <tt>CancerEngine</tt>.
```{r}
ce <- ClinicalEngine(20, 4, isWeighted = TRUE)
summary(ce)
```
Note that the prevalences are not equal; when you use <tt>isweighted = TRUE</tt>,
they are chosen from a Dirichlet distribution.
Note also that the <tt>summary</tt> function describes the object as a 
<tt>CancerEngine</tt>, since the same underlying structure is used to
implement a <tt>ClinicalEngine</tt>.

Now we confirm that the model expects to produce the 20 features that we
requested. It will do so using 10 "components", where each component consists
of a pair of correlated features.
```{r nrow}
nrow(ce)
nComponents(ce)
```

## Simulate Raw Data
The <tt>ClinicalEngine</tt> is used to simulate the raw, base dataset.
```{r}
dset <- rand(ce, 300)
```

Data are simulated as a list with two objects: simulated 
<tt>data</tt> and associated <tt>clinical</tt> information, including 
"ground truth" subtype membership and survival data (outcome, length of 
followup, and occurrence of event of interest within the followup period).
```{r}
class(dset)
names(dset)
summary(dset$clinical)
```

The raw <tt>data</tt> are simulated as a matrix of continuous values.
```{r}
class(dset$data)
dim(dset$data)
```

## Apply Clinically Realistic Noise
The user may add further additive noise to the raw data. The 
<tt>ClinicalNoiseModel</tt> simulates additive noise for each 
feature _f_ and patient _i_ as a normal distribution 
$E_{fi} \sim N(0, \tau)$ , where the standard deviation $\tau$ 
varies with a hyperparameter along the gamma distribution 
$\tau \sim Gamma(shape, scale)$. Thus, the ClinicalNoiseModel 
generates many features with low noise (such as a tightly calibrated 
laboratory test) and some features with high noise (such as 
a blood pressure measured by hand and manually entered into the 
medical record.) The user may apply default parameters or individual 
parameters. Next, the <tt>ClinicalNoiseModel</tt> is applied to 
<tt>blur</tt> the previously simulated data. The default model 
below generates a low overall level of additive noise.

```{r}
cnm <- ClinicalNoiseModel(nrow(ce@localenv$eng), shape = 1.02, scale = 0.05)
summary(cnm)
noisy <- blur(cnm, dset$data)
```

## Simulate Mixed-Type Data
<tt>Umpire 2.0</tt> allows the simulation of binary, nominal, 
and ordinal data from raw, continuous data in variable, user-defined 
mixtures. The user defines prevalences, summing to 1, of binary, 
continuous, and categorical data in the desired final mixture.
For categorical features, the user may tune the percent of categorical 
data desired to be nominal and the range of the number of categories to be 
simulated. 

The data simulated above by the <tt>ClinicalEngine</tt> and 
<tt>ClinicalNoiseModel</tt> takes rows (not columns) as features, as 
an omics convention. Thus, by default, when generating data,
rows are treated as features and columns as patients. The <tt>makeDataTypes</tt>
method transposes its results to a data frame where the columns are features
and the rows are patients. This transposition both fits better with the
conventions used for clinical data, but also supports the ability to store
different kinds of (mixed-type) data in different columns.

```{r}
dt <- makeDataTypes(dset$data,
                   pCont = 1/3, pBin = 1/3, pCat = 1/3,
                   pNominal = 0.5, range = 3:9,
                   inputRowsAreFeatures = TRUE)
names(dt)
```

The <tt>makeDataTypes</tt> function generates a list containing two objects:
a data.frame of mixed-type data...
```{r}
class(dt$binned)
dim(dt$binned)
summary(dt$binned)
```

The <tt>cutpoints</tt> contain a record, for each feature, of data 
type, break points, and labels.  Here are two examples of the kind of
information stored for a cutpoint.
```{r}
dt$cutpoints[[1]]
dt$cutpoints[[5]]
```
And here is an overview of the number of features of each type.
```{r}
cp <- dt$cutpoints
type <- sapply(cp, function(X) { X$Type })
table(type)
```

The <tt>cupoitns</tt> should be saved for downstream use in 
the <tt>MixedTypeEngine</tt>.


## The MixedTypeEngine
The many parameters defining a simulated data mixture can be stored as 
a single <tt>MixedTypeEngine</tt> for downstream use to easily generate 
future datasets with the same simulation parameters.

The <tt>MixedTypeEngine</tt> stores the following components for re-implementation:

1. The <tt>ClinicalEngine</tt>, including parameters for generating the subtype pattern and survival model.
2. The <tt>ClinicalNoiseModel</tt>.
3. The <tt>cutpoints</tt> generated by <tt>makeDataTypes</tt>.

```{r}
mte <- MixedTypeEngine(ce,
                       noise = cnm,
                       cutpoints = dt$cutpoints)
summary(mte)
```

With <tt>rand</tt>, the user can easily generate new data sets with 
the same simulation parameters.
```{r}
dset2 <- rand(mte, 20)
class(dset2)
summary(dset2$data)
summary(dset2$clinical)
```

By using the <tt>keepal</tt> argument othe function, you can keep the
intermediate datasets produced by the <tt>rand</tt> method.
```{r}
dset3 <- rand(mte, 25, keepall = TRUE)
class(dset3)
names(dset3)
```
The <tt>raw</tt> and <tt>noisy</tt> elements have the rows as (future clinical)
features and the columns as patients/samples.
```{r raw}
dim(dset3$raw)
summary(t(dset3$raw))
dim(t(dset3$noisy))
summary(dset3$noisy)
```
Noisy data arises by adding simulated noise to the raw data.
```{r, fig.cap="Raw and noisy data."}
plot(dset3$raw[5,], dset3$noisy[5,], xlab = "Raw", ylab = "Noisy", pch=16)
```

The <tt>binned</tt> element has columns as features and rows as samples.
Binned data arises by applying cut points to noisy data.
```{r fig.cap = "Noisy and binned data."}
dim(dset3$binned)
summary(dset3$binned)
plot(dset3$binned[,5], dset3$noisy[5,], xlab = "Binned", ylab = "Noisy")
```

