---
title: "Umpire 2.0: Clinically Realistic Simulations"
author: "Kevin R. Coombes and Caitlin E. Coombes"
data: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Umpire}
  %\VignetteKeywords{Umpire}
  %\VignetteDepends{Umpire}
  %\VignettePackage{Umpire}
  %\VignetteEngine{knitr::rmarkdown}
---

## Introduction

Version 2.0 of the Ultimate Microarray Prediction, Inference, and 
Reality Engine (Umpire) extends the functions of the Umpire 1.0 R 
package to allow researchers to simulate realistic, mixed-type, 
clinical data. Statisticians, computer scientists, and clinical 
informaticians who develop and improve methods to analyze clinical data 
from a variety of contexts (including clinical trials, population 
cohorts, and electronic medical record sources) recognize that it is 
difficult to evaluate methods on real data where "ground truth" is unknown. 
Frequently, they turn to simulations where the can control the 
underlying structure, which can result in simulations which are too 
simplistic to reflect complex clinical data realities. Clinical 
measurements on patients may be treated as independent, in spite of 
the elaborate correlation structures that arise in networks, pathways, 
organ systems, and syndromes in real biology. Further, the researcher 
finds limited tools at her disposal to facilitate simulation of binary, 
categorical, or mixed data at this representative level of biological \
complexity.


In this vignette, we describe a workflow with the <tt>Umpire</tt> package 
to simulate biologically realistic, mixed-type clinical data.

As usual, we start by loading the package:

```{r}
library(Umpire)
library(Polychrome)
```

## Simulating Mixed-Type Clinical Data

The simulation workflow begins by simulating complex, correlated, 
continuous data with known "ground truth" by implementing a 
<tt>ClinicalEngine</tt>. We simulate 20 features and 4 clusters of 
unequal size.
```{r}
ce <- Umpire:::ClinicalEngine(20, 4, isWeighted=TRUE)

```



## The MixedTypeEngine


















Next, we can look at one of the color palettes it provides.

```{r}
mypal <- kelly.colors(22)
```

```{r fig.width=7, fig.height=5}
swatch(mypal)
```

## Resorted barplots

To get a better idea of whether colors are distinguishable, we can
sort them by hue:
```{r fig.width=7, fig.height=5}
swatchHue(mypal)
```

or by luminance:
```{r fig.width=7, fig.height=5}
swatchLuminance(mypal)
```

or in random order:
```{r fig.width=7, fig.height=5}
ranswatch(mypal)
```

## Scatter Plots
Some colors that work well in barplots may work less well in scatter
plots.  To test that possibility, we can generate some random points
and plot them. The next line of code generates 14 points from each
color in the "alphabet" palette.

```{r fig.width=7, fig.height=5}
pal2 <- alphabet.colors(26)
ranpoints(pal2, 14)
```

Our standard way to think about colors is in the L\*u\*v\* color space
model defined by the CIE.  The next plot shows how the points in the
palette are distributed in the u-v subspace.
```{r fig.width=7, fig.height=5}
uvscatter(pal2)
```

We can also show the associated luminance, L.
```{r fig.width=7, fig.height=5}
luminance(pal2)
```

## Plotting Curves
We can also test if a palette is useful for drawing a graph full of
different curves rarther than points.
```{r fig.width=7, fig.height=5}
rancurves(pal2)
```

## Color Similarity
We use L\*u\*v\* color space because perceptual distinguishability seems
to be closely related to Euclidean distance in this space.  To get
another view of which colors are most similar, we can perform
hierarchical clustering of colors using this representation.
```{r fig.width=7, fig.height=5}
plothc(pal2)
```

We can also perform principal components in L\*u\*v\* space to display
a projection that tries to preserve distances.
```{r fig.width=7, fig.height=5}
plotpc(pal2)
```

## Grayscale
One further consideration that should be kept in mind when creating
color palettes is that not every potential viewer has normal color
vision. One way to test if a color palette will remain useful in this
context is to convert it to gray scale. We do this by converting to
grayscale using the desaturate function from the colorspace package,
and then preparing a barplot by luminance. 

```{r fig.width=7, fig.height=5}
grayed <- colorspace::desaturate(pal2)
swatchLuminance(grayed)
```
